### 2012-09-24 hdf5/discs_2012/abstract.tex
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

HDF5 is a data model, library and file format for storing and managing data. It is designed for flexible and efficient I/O for high volume and complex data. Natively, it uses a single-file format where multiple HDF5 objects are stored in a single file. In a parallel HDF5 application, multiple processes access a single file, thereby resulting in a performance bottleneck in I/O. Additionally, a single-file format does not allow semantic post processing on individual objects outside the scope of the HDF5 application. We have developed a new plugin for HDF5 using its Virtual Object Layer that provides serves two purposes: 1) it uses PLFS to convert the single-file layout into a data layout that is optimized for the underlying file system, and 2) it stores data in a unique way that enables semantic post-processing on the data. We measure the performance of the plugin and discuss work leveraging the new semantic post-processing functionality enabled.  We further discuss the applicability of this approach for exascale burst buffer storage systems.

>>>>

HDF5 is a data model, library and file format for storing and managing data. It is designed for flexible and efficient I/O for high volume and complex data. Natively, it uses a single-file format where multiple HDF5 objects are stored in a single file. In a parallel HDF5 application, multiple processes access a single file, thereby resulting in a performance bottleneck in I/O. Additionally, a single-file format does not allow semantic post processing on individual objects outside the scope of the HDF5 application. We have developed a new plugin for HDF5 using its Virtual Object Layer that serves two purposes: 1) it uses PLFS to convert the single-file layout into a data layout that is optimized for the underlying file system, and 2) it stores data in a unique way that enables semantic post-processing on data. We measure the performance of the plugin and discuss work leveraging the new semantic post-processing functionality enabled.  We further discuss the applicability of this approach for exascale burst buffer storage systems.

### 2012-09-24 hdf5/discs_2012/background.tex
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

\subsection{HDF5}
HDF5 is a versatile data model containing complex data objects and metadata. 
Its information set is a collection of datasets, groups, datatypes and metadata objects. The data model defines mechanisms for creating associations between various information items. The main components of HDF5 are described below.

File:
In the HDF5 data model the container of an HDF5 infoset
is represented by a file. It is a collection of objects and also explains the relationship between them.
Every file begins with a root group "/", which serves as the "starting-point" in the object hierarchy.

@@

Attribute:
Attributes are used for annotating datasets, groups, and datatype objects. They are datasets themselves, and are \textit{attached} to the existing objects they annotate.

@@

Virtual Object Layer (VOL):
The Virtual Object Layer (VOL) is a new abstraction layer internal to the HDF5 library~\cite{vol}. It is implemented just below the public API. The VOL exports an interface that allows writing plugins for HDF5, thereby enabling developers to store objects in a format different from the default HDF5 file format (like native netCDF or HDF4 format). Plugin writers provide an implementation for a set of functions that access data on disk. These include functions for file management, dataset creation and access, group creation, to name a few.

\subsection{PLFS}
PLFS is a middleware virtual file system that converts writes to a shared logical file into writes to multiple physical files. 
It is situated between the application and the parallel file system responsible for the actual data storage. 
It transforms N-1 into N-N, where every process participating in I/O writes data to its own, separate file. 
The basic operation of PLFS is as follows. For every writer to a logical file, PLFS creates a unique physical file on the underlying parallel file system. 
It also maintains sufficient metadata to recreate the shared logical file. 
We added a new feature to PLFS called extendible attributes (Xattrs). Xattrs serve as short, extensible metadata stored as key-value pairs.
They can be used to store user-defined information about data for easy and fast retrieval. 
%We use xattrs to store metadata about HDF5 datasets. For example, the datatype, number of dimensions and their extent are stored as xattrs.

Users can interface with PLFS directly by using the PLFS API or by using an MPI library that has been integrated with PLFS (ad\_plfs layer).

>>>>

%\subsection{HDF5}
HDF5 is a versatile data model containing complex data objects and metadata. 
Its information set is a collection of datasets, groups, datatypes and metadata objects. The data model defines mechanisms for creating associations between various information items. The main components of HDF5 are described below.

File:
In the HDF5 data model the container of an HDF5 infoset
is represented by a file. It is a collection of objects that also explains the relationship between them.
Every file begins with a root group "/", which serves as the "starting-point" in the object hierarchy.

@@

Attribute:
Attributes are used for annotating datasets, groups, and datatype objects. They are datasets themselves, and are \textit{attached} to existing objects they annotate.

@@

Virtual Object Layer (VOL):
VOL is a new abstraction layer internal to the HDF5 library~\cite{vol}. It is implemented just below the public API. The VOL exports an interface that allows writing plugins for HDF5, thereby enabling developers to store objects in a format different from the default HDF5 file format (like native netCDF or HDF4 format). Plugin writers provide an implementation for a set of functions that access data on disk. These include functions for file management, dataset creation and access, group creation, to name a few.

%\subsection{PLFS}
PLFS is a middleware virtual file system that converts writes to a shared logical file into writes to multiple physical files. 
It is situated between the application and the parallel file system responsible for the actual data storage. 
It transforms N-1 into N-N, where every process participating in I/O writes data to its own, separate file. 
The basic operation of PLFS is as follows. For every writer to a logical file, PLFS creates a unique physical file on the underlying parallel file system. 
It also maintains sufficient metadata to recreate the shared logical file. 
We added a new feature to PLFS called extendible attributes (Xattrs). Xattrs serve as short, extensible metadata stored as key-value pairs.
They can be used to store user-defined information about data for easy and fast retrieval. 
%We use xattrs to store metadata about HDF5 datasets. For example, the datatype, number of dimensions and their extent are stored as xattrs.

Users can interface with PLFS directly by using the PLFS API or by using its MPI-IO driver (ad\_plfs).

### 2012-09-24 hdf5/discs_2012/conclusion.tex
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

Using HDF5's virtual object layer, we have developed a new plugin that makes use of PLFS for improved I/O performance. We deviate from HDF5's native file format and store data in a way that allows us to perform useful post-processing independently on HDF5 objects. 
Such a format can further allow us to incorporate semantic analysis of data for burst buffer type workloads. 
Initial results show that our plugin consistently performs better than regular MPI-IO, whereas MPI-IO when used with PLFS's MPI-IO driver shows the best results.

>>>>

Using HDF5's virtual object layer, we have developed a new plugin that makes use of PLFS for improved I/O performance. We deviate from HDF5's native file format and store data in a way that allows us to perform useful post-processing independently on HDF5 objects. 
Such a format can further allow us to incorporate semantic analysis of data for burst buffer type workloads. 
Initial results show that our plugin outperforms MPI-IO in most cases, whereas MPI-IO when used with PLFS's MPI-IO driver shows the best results.

### 2012-09-24 hdf5/discs_2012/evaluation.tex
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

For evaluation purposes, we have used HDF5's \textit{h5perf} performance tool~\cite{h5perf}. It allows configuring various parameters, such as the number of processes to run the test with, number of datasets to be created in the file, amount of data contributed by each process per dataset, amount of data read/written by a process in a single I/O call (transfer size) etc. For our measurements, we created 10 datasets in the file, and the total file size was 64 GB or more.

All tests were performed on the Lustre parallel file system~\cite{lustre} on the Atlas cluster at University of Dresden. The file system has 12 OSTs with a stripe size of 1MB. The file system is connected to the compute nodes using an SDR Infiniband link. It should be noted that the file system is in production and subject to demanding and varying workload. We present here the average bandwidth values of 3 runs, though in certain cases, we have seen significant variation.
The cluster has 92 AMD Opteron nodes with 64 cores each and 64 to 512 GB memory.

We have performed tests with 1,2,4,8,32 and 64 processes. We spawned a maximum of 4 processes per node.
Reads and writes are either contiguous or interleaved; processes either access contiguous locations in file or execute a strided pattern.
We compare the performance of the default MPI-IO driver, our plugin, and the PLFS MPI-IO driver (ad\_plfs).

In Figure~\ref{write_contig}, we show the write performance for a transfer size of 1MB for contiguous writes, for varying number of processes. It can be seen that the plugin regularly outperforms MPI-IO. However, as we increase the number of processes, its performance falls below that of ad\_plfs. Figure~\ref{read_contig} shows the performance of contiguous reads for an unaligned transfer size. The plugin again outperforms MPI-IO, but the performance of ad\_plfs is the best for 64 processes. 
%Figure~\ref{write_interleaved} shows the performance of unaligned writes for a maximum of 8 processes. We set the transfer size here to be 10 bytes more than 1MB. Also, the writes here are interleaved as opposed to contiguous. We see a similar pattern, where the plugin outperforms MPI-IO, but is not as good as that of ad\_plfs for 8 processes. 
Figure~\ref{write_collective} compares the performance of MPI-IO with that of the plugin for collective, unaligned, interleaved writes for a maximum of 8 processes. It is important to note that the plugin outperforms MPI-IO in collective mode. We have not been able to obtain results for collective ad\_plfs due to resource restrictions.

Overall, our results show that although the plugin generally shows good performance, it does not scale as well as we increase the number of processes. This is due to the fact that since we are directly making calls to the PLFS API, the metadata overhead seems to be high. We have all processes participating in file open and close operations, which for large problem sizes can add significant overhead. Hence, for future work, we plan to use ad\_plfs directly in the plugin which can overcome this performance drawback, since ad\_plfs has collective optimizations in open and close. It should however be noted that there are some MPI libraries tailored to suit a specific purpose and which do not provide an implementation for MPI-IO. Such libraries can benefit from using a plugin that does not rely on MPI-IO.

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4PEs_node_1M_contig_ind_writes}
\caption{Performance of contiguous writes}
\label{write_contig}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4pes_node_1m_ind_read}
\caption{Performance of contiguous reads}
\label{read_contig}
\end{figure}

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{interleaved_1M10_4pes_writes_noncoll}
%\caption{Performance of interleaved writes}
%\label{write_interleaved}
%\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{4pes_interleaved_1m10_coll_reads}
\caption{Performance of collective writes}
\label{write_collective}
\end{figure}

>>>>

For evaluation purposes, we have used HDF5's \textit{h5perf} performance tool~\cite{h5perf}. It allows configuring various parameters, such as the number of processes, number of datasets, amount of data read/written by a process in a single I/O call (transfer size) etc. For our measurements, we created 10 datasets, and the total file size was 64 GB or more.
In every run, every process contributed equal amount of data per dataset using the default individual (non-collective) I/O mode of h5perf.

Tests were performed on the Lustre parallel file system~\cite{lustre} on the Atlas cluster at University of Dresden. The file system has 12 OSTs with a stripe size of 1MB. The file system is connected to the compute nodes using an SDR Infiniband link. The cluster has 92 AMD Opteron nodes with 64 cores each and 64 to 512 GB memory. Tests were run thrice and we present the average bandwidth values, which does not include the time taken to open and close the file.

We have performed tests with 1,2,4,8,32 and 64 processes with a maximum of 4 processes per node.
Reads and writes are either contiguous or interleaved; processes either access contiguous locations in file or execute a strided pattern.
We compare the performance of the default MPI-IO driver, our plugin, and the PLFS MPI-IO driver (ad\_plfs).

In Figure~\ref{write_contig}, we show the write performance for a transfer size of 1MB for contiguous writes. It can be seen that the plugin regularly outperforms MPI-IO except for the 64 process case, where the metadata overhead incurred by the plugin is high. The performance  of ad\_plfs is the best for higher process counts. Figure~\ref{read_contig} shows the performance of contiguous reads. 
%Figure~\ref{write_interleaved} shows the performance of unaligned writes for a maximum of 8 processes. We set the transfer size here to be 10 bytes more than 1MB. Also, the writes here are interleaved as opposed to contiguous. We see a similar pattern, where the plugin outperforms MPI-IO, but is not as good as that of ad\_plfs for 8 processes. 
Figures~\ref{write_interleaved} and~\ref{read_interleaved} show the interleaved write and read performance respectively for an unaligned transfer size (1M + 10bytes) for a maximum of 8 processes. The write performance of MPI-IO is quite poor in this case. The plugin easily outperforms MPI-IO and almost matches the performance of ad\_plfs.

Overall, results show that the plugin consistently shows good performance, however it does not scale as well as ad\_plfs. This is due to the fact that since we are directly making calls to the PLFS API, the metadata overhead is high. All processes participate in file open and close operations, which for large problem sizes adds significant overhead. Hence, for future work, we plan to use ad\_plfs directly in the plugin which can overcome this performance drawback, since ad\_plfs has collective optimizations in open and close. It should be noted that there are some MPI libraries tailored to suit specific types of applications and which do not provide an implementation for MPI-IO. Such libraries can benefit from using a plugin that does not rely on MPI-IO.

\begin{figure}[!t]
\centering
\includegraphics[width=3.6in,height=3.0in]{contig_w}
\caption{Performance of contiguous writes}
\label{write_contig}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=3.6in,height=3.0in]{contig_r}
\caption{Performance of contiguous reads}
\label{read_contig}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=3.6in,height=3.0in]{interleaved_w}
\caption{Performance of interleaved, unaligned writes}
\label{write_interleaved}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=3.6in,height=3.0in]{interleaved_r}
\caption{Performance of interleaved,unaligned reads}
\label{read_interleaved}
\end{figure}

### 2012-09-24 hdf5/discs_2012/introduction.tex
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

Hierarchical Data Format (HDF5) is a technology suite for efficient management of large and complex data~\cite{hdf5_ad11}. It supports complex relationships between data and dependencies between objects. HDF5 is widely used in industry and scientic domains, in understanding global climate change, special effects in film production, DNA analysis, weather prediction, financial data management etc.~\cite{hdf5_users}
Parallel HDF5 (PHDF5)~\cite{phdf5} enables developing high performance, parallel applications using standard technologies like MPI in conjunction with HDF5. 
%An HDF5 file created using PHDF5 is compatible with serial HDF5 files and is shareable between different platforms. 
PHDF5 exports a standard parallel I/O interface which itself uses MPI's parallel I/O functionality. This is used alongwith parallel file systems to achieve high performance I/O.

An HDF5 file is a self-describing format which combines data and metadata. It is a container in which users typically store multiple HDF5 objects alongside their metadata. 
However, this native single-file format has its disadvantages. A PHDF5 application has multiple processes accessing a single file (i.e. N-1 access pattern). Unfortunately, many popular parallel file systems are known to behave poorly under these circumstances~\cite{lustre_perf}\cite{nersc_io}. Secondly, since many HDF5 objects are stored in a single file, this complicates the possibility of performing useful semantic operations on individual objects. In this paper, we enhance the performance of HDF5 by addressing the performance issues inherently arising from poor file system performance on N-1 access patterns, and augment performing  useful post-processing on HDF5 objects.

@@

In this work, we have developed a new plugin for HDF5 using its recently introduced Virtual Object Layer (VOL).  
This plugin serves two purposes: it stores data in a unique way that enables semantic post-processing on HDF5 objects, and by using PLFS to convert N-1 accesses into N-N accesses, thereby resulting in improved I/O performance.

>>>>

Hierarchical Data Format (HDF5) is a technology suite for efficient management of large and complex data~\cite{hdf5_ad11}. It supports complex relationships between data and dependencies between objects. HDF5 is widely used in industry and scientific domains, in understanding global climate change, special effects in film production, DNA analysis, weather prediction, financial data management etc.~\cite{hdf5_users}
Parallel HDF5 (PHDF5)~\cite{phdf5} enables developing high performance, parallel applications using standard technologies like MPI in conjunction with HDF5. 
%An HDF5 file created using PHDF5 is compatible with serial HDF5 files and is shareable between different platforms. 
PHDF5 exports a standard parallel I/O interface which itself uses MPI's parallel I/O functionality. This is used along with parallel file systems to achieve high performance I/O.

An HDF5 file is a self-describing format which combines data and metadata. It is a container in which users typically store multiple HDF5 objects alongside their metadata. 
However, this native single-file format has its disadvantages. A PHDF5 application has multiple processes accessing a single file (i.e. N-1 access pattern). Unfortunately, many popular parallel file systems are known to behave poorly under these circumstances~\cite{lustre_perf,nersc_io}. Secondly, since many HDF5 objects are stored in a single file, this complicates performing useful semantic operations on individual objects.

@@

In this paper, we enhance the performance of HDF5 by addressing the performance issues inherently arising from poor file system performance on N-1 access patterns, and augment performing  useful post-processing on HDF5 objects.
We have developed a new plugin for HDF5 using its recently introduced Virtual Object Layer (VOL).  
This plugin stores data in a unique way that enables semantic post-processing on HDF5 objects, and uses PLFS to convert N-1 accesses into N-N accesses, thereby resulting in improved I/O performance.

### 2012-09-24 hdf5/discs_2012/plugin.tex
A couple of typos found by running aspell on the tex files
Final submitted version on 9/22. Evaluation contains plots from old 64 filesize test case. Minor typos, bug fixes everywhere. Removed old figures and added new ones.

We have developed a plugin which stores data in a format which enables performing semantic analysis on the data and uses PLFS for high I/O throughput. 
Instead of storing all objects in a single file, it stores every HDF5 object in a separate location, so that data from different objects is not contained in the same file. In short, we provide a raw mapping of the HDF5 objects to the file system. 
HDF5 files and groups are stored as directories, whereas datasets and attributes, which contain raw data, are stored in files created using PLFS. 
Consider the example shown in figure~\ref{hdf5_example}. Using the plugin, file Sample.h5 is stored as a directory. Group G1 is stored as a directory under it, and datasets D1 and D2 are files. Attribute A1 is a file created at the same location as dataset D1 to which it is attached. Additionally, the name of the attribute is stored as \textit{dataset name}.\textit{attribute name} to denote the dataset to which the attribute is attached. Thus these objects are stored at the following paths:\\
/Sample.h5/\\
/Sample.h5/G1\\
/Sample.h5/G1/D3\\
/Sample.h5/D1\\
/Sample.h5/D1.A1\\
/Sample.h5/D2\\

We can see that the relationship between the objects is represented by their relative paths at the file system. That is, the path /Sample.h5/G1/D3 tells us that D3 is a dataset belonging to Group G1 under the root group of the file. This way, the plugin also eliminates the need to store the metadata describing the relationship between objects. Metadata about datasets, such as the datatype, extent, dimensions etc. are stored as Xattrs. 
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{hdf5_example_plugin}
%\caption{HDF5 file created using our PLFS plugin}
%\label{hdf5_example_plugin}
%\end{figure}

This approach gives us the ability to distinguish HDF5 objects inside the storage system.

Our plugin makes direct calls to the PLFS API. For the current version of our plugin, we do not support collective I/O operations. It should be noted that HDF5 when used natively with MPI-IO allows users to specify whether collective I/O should be used for reading and writing data.

The above format of storing data allows us to perform atleast two different analysis optimizations.  To illustrate via an example, imagine storing a three-dimensional ocean model within a PLFS file. The storage system sees the file as an opaque linear array of bytes.  With the structure, however, PLFS can provide \textit{active analysis} as well as \textit{semantic restructuring}.

Active analysis borrows the transducers idea from the Semantic File System~\cite{semantic_fs} which has since been productized in BigTable and Hbase~\cite{google_coprocessors,GFS,apache_hbase}.  With active analysis, the application can ship a data parser function when it creates the PLFS file.  As the data is written into PLFS, PLFS can apply the data function on the streaming data.  The function will output key-value pairs which PLFS can embed in its extensible metadata.  In this example, one simple function might record the height of the largest wave.  Due to PLFS's model of storing a logical file across multiple physical files, the PLFS extensible metadata can record the height of the largest wave within each physical file.  However, given that PLFS now understands the structure of the logical file, these multiple physical files, within the PLFS container, are more accurately thought of as shards.  In a future burst buffer architecture~\cite{burst_buffers}, these semantic shards will be spread across multiple burst buffer nodes.  Therefore, subsequent analysis of the ocean model can quickly find the burst buffer containing the shard with the largest wave by searching a small amount of extensible metadata instead of scanning the entire ocean model.

>>>>

We have developed a plugin which stores data in a format which enables performing semantic analysis on the data and uses PLFS for high I/O throughput. 
Instead of storing all objects in a single file, it stores every HDF5 object in a separate location, so that data from different objects is not contained in the same file. In short, we provide a raw mapping of HDF5 objects to the file system. 
HDF5 files and groups are stored as directories, whereas datasets and attributes, which contain raw data, are stored in files created using PLFS. 
Consider the example shown in figure~\ref{hdf5_example}. Using the plugin, file Sample.h5 is stored as a directory. Group G1 is stored as a directory under it, and datasets D1 and D2 are files. Attribute A1 is a file created at the same location as dataset D1 to which it is attached. Additionally, the name of the attribute is stored as \textit{dataset name}.\textit{attribute name} to denote the dataset to which the attribute is attached. Thus these objects are stored at the following paths:\\
/Sample.h5/\\
/Sample.h5/G1\\
/Sample.h5/G1/D3\\
/Sample.h5/D1\\
/Sample.h5/D1.A1\\
/Sample.h5/D2\\

We can see that the relationship between the objects is represented by their relative paths at the file system. That is, the path /Sample.h5/G1/D3 tells us that D3 is a dataset belonging to Group G1 under the root group of the file. This approach gives us the ability to distinguish HDF5 objects inside the storage system. Also, the plugin eliminates the need to explicitly store the metadata describing the relationship between objects. Metadata about datasets, such as the datatype, extent, dimensions etc. are stored as Xattrs.

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{hdf5_example_plugin}
%\caption{HDF5 file created using our PLFS plugin}
%\label{hdf5_example_plugin}
%\end{figure}

Our plugin makes direct calls to the PLFS API. For the current version, we do not support collective I/O operations. It should be noted that HDF5 when used natively with MPI-IO allows users to specify whether collective I/O should be used for reading and writing data.

The above format of storing data allows us to perform at least two different analysis optimizations.  To illustrate via an example, imagine storing a three-dimensional ocean model within a PLFS file. The storage system sees the file as an opaque linear array of bytes.  With the structure, however, PLFS can provide \textit{active analysis} as well as \textit{semantic restructuring}.

Active analysis borrows the transducers idea from the Semantic File System~\cite{semantic_fs} which has since been productized in Google's BigTable and Apache Hbase technologies~\cite{google_coprocessors,GFS,apache_hbase}.  With active analysis, the application can ship a data parser function when it creates the PLFS file.  As the data is written into PLFS, PLFS can apply the data function on the streaming data.  The function will output key-value pairs which PLFS can embed in its extensible metadata.  In this example, one simple function might record the height of the largest wave.  Due to PLFS's model of storing a logical file across multiple physical files, the PLFS extensible metadata can record the height of the largest wave within each physical file.  However, given that PLFS now understands the structure of the logical file, these multiple physical files, within the PLFS container, are more accurately thought of as shards.  In a future burst buffer architecture~\cite{burst_buffers}, these semantic shards will be spread across multiple burst buffer nodes.  Therefore, subsequent analysis of the ocean model can quickly find the burst buffer containing the shard with the largest wave by searching a small amount of extensible metadata instead of scanning the entire ocean model.
